Addressing climate change and development pressures in an urban estuary through habitat restoration planning

### 2022-08-24 manu-draft.qmd
intro revision

The health of estuarine systems, coastal habitats, and associated fauna and flora are inextricably linked to land uses and management throughout the watershed [@Yoskowitz15] These habitats provide multiple ecosystem services, including wildlife shelter and migratory corridors [@Yoskowitz15], fisheries production [@Houde93], water quality improvement [@Kushlan90; @Sprandel00; @AvilaGarcia20], erosion and flood attenuation [@Calil15; @Menendez18], carbon sequestration [@Dontis20] and recreation [@Chung18]. The Tampa Bay Estuary Program (TBEP) is one of 28 programs administered by the US Environmental Protection Agency (USEPA) under the National Estuary Program (NEP). In recognition of threats to habitats from development and climate change stressors, the TBEP and partners recently created the third iteration of a plan to establish targets and goals for habitat restoration within the Tampa Bay watershed. As an NEP, the program has guided regional environmental restoration initiatives for the estuary since 1991. The methodologies used in the creation of a 2020 Habitat Master Plan Update for Tampa Bay [@Robison20] are highly transferable to both coastal and non-coastal systems.

## Development, climate change, and coastal squeeze

Development causes multiple perturbations within any watershed. The accumulated impacts of construction and associated infrastructure remove or substantially modify existing habitats, and can alter hydrology of nearby streams and rivers [@Theobald97; @Sim14]. Given the degree to which the Tampa Bay watershed has been urbanized, the synergistic effects of continued coastal development and future climate change are primary concerns for maintenance of estuarine and coastal habitat health. Observed and potential adverse effects of climate change and sea level rise on marine and estuarine ecosystems are well-documented [@Scavia02; @Spalding07; @Titus09]. With regard to estuarine habitats, the primary concerns are that sea level rise is now occurring at such a rapid rate that the landward migration of tidal wetlands in response cannot keep pace; or that the upland slope has already been lost to urban development and hardening, leaving no place for tidal wetland migration [@Titus09]. Geological, physical and chemical changes could include alterations in sediment deposition and erosion patterns, micro-topography, and water quality [@Arnell15; @Whitehead09].

Related changes in habitats in response to climate change (sea level rise and warming) include landward migration of mangroves into salt marshes, upstream migration of salt marshes within tidal tributaries, and upland forest migration [@Brinson95; @Vogelmann12; @Cavanaugh19]. Species dependent upon these habitats will be forced to change use patterns or adapt to the new conditions [@Iwamura13]. For example, while black needle rush (*Juncus roemerianus*) can tolerate a wide salinity range [@Eleuterius79; @Stout84], the largest remaining *J. roemerianus* marshes in Tampa Bay are located in the lower-salinity reaches of tidal rivers and creeks. The greatest extents of those marshes occur in river systems where their upstream extent is constrained by impoundments for public water supplies. Spatial restriction in these hydrologically truncated rivers may make these marshes particularly vulnerable to "pinching out", as upstream migration in response to sea level rise will be cut off by anthropogenic barriers. Similarly, landward migration of salt barrens (high marsh areas in Tampa Bay) in response to sea level rise will be restricted by the filling and hardening of coastal uplands associated with existing or future urban development.

## Need for a paradigm shift

The synergistic effects of development and climate change diminish the available space for future restoration in urbanizing estuaries, thereby impacting the variety of ecosystem services provided by these habitats and the wildlife they support [@Enwright15]. Given projected habitat losses without intervention efforts and the limited resources (including time, land, funding, and labor force) available, it is important to appropriately and realistically site restoration projects to increase the likelihood of success. To achieve this objective, a new restoration approach on a broad watershed scale will be implemented.

Previously, a "retrospective" approach to setting habitat protection and restoration targets in Tampa Bay was employed [@Lewis96; @Robison10; @Cicchetti11; @Russell15]. Under this paradigm, priority was given to restoration activities focused on habitat types, important for a suite of ten estuarine faunal guilds, that were disproportionately lost or degraded compared to a benchmark period. Primary criticisms of this approach included a lack of consideration for future sea level rise and other climate change factors [@Yoskowitz15], use of expanded and different habitats outside the Tampa Bay watershed by the faunal guilds [@Robison10], lack of attention to upland or freshwater wetland habitats, and little recognition of other stressors such as land development trends or actual available space for restoration efforts.

Past approaches for guiding restoration planning have been successfully used in other contexts, but they do not fully balance competing needs. For example, an integrated watershed approach [@EPA96] has been utilized since the early 1990s to diagnose and manage water quantity and quality problems that have contributed to seagrass restoration in the system. Additionally, the habitat mosaic approach [@Henningsen05] of including multiple habitat types within restoration projects is recognized as necessary in Tampa Bay [@Hughes11] and elsewhere to allow for ecosystem state changes in response to different environmental pressures [@Duarte09; @Palmer09].

Adaptive management [@Holling78; @Gregory06] components have been increasingly used to address challenges of sea level rise, climate change, and development stressors, including monitoring to identify critical restoration decision points and needed intervention with contingency plans. Rising sea levels and temperatures and altered rainfall patterns are causing observable changes to habitats on a global scale [@Cavanaugh14; @Garner15; @Yoskowitz15], including within Tampa Bay [@Price17], and those changes are expected to become more pronounced over the next several decades [@Sheehan16; @Nerem18].

## Updating the approach

The new approach integrates the whole watershed, addresses historical changes, focuses on trajectories that have occurred during more contemporary time periods, and considers both current and future stressors -- particularly land development and sea level rise. There is relatively consistent extent and distribution data for most Tampa Bay habitats of interest (1988 to 2018), representing a time period when federal, state and local regulations were in effect and regional impacts from climate change are documented [@Raabe12; @Cavanaugh14]. This approach establishes a broader framework that guides both watershed-level habitat master planning and site-level restoration design activities and incorporates applicable elements of the other habitat restoration paradigms discussed above [@Palmer09]. Our broader framework for guiding restoration activities includes 1) designation of habitat types by strata relative to the aquatic-terrestrial gradient, 2) quantification of historical trends by habitat types to identify appropriate future targets in acreage, and 3) identification of opportunity areas that could be used by practitioners to achieve restoration goals based on habitat type and past trajectories.

With regard to habitat restoration projects, the design approach must envision not only what is possible today, but also what the coastal landscape will look like in 50 years and beyond. Design features should continue to use the historical "habitat mosaic" approach, but should also include coastal upland features that accommodate tidal inundation and the landward advance of emergent tidal wetlands [@Enwright16].

@@

Addressing the full suite of habitats in a watershed is now recognized as critical for large-scale restoration planning efforts [@Palmer09; @Lamb18]**.** The major habitat types of Tampa Bay can be described and organized relative to tidal influence and location in the watershed. Subtidal habitats include those that are submerged all or most of the time; emergent tidal wetlands include those that are submerged during high tides but exposed during low tides; and supratidal habitats include those that occur above the high tide line.

@@

Three strata, distinct geographic breakpoints in the watershed where different habitat management and restoration activities could take place, were created to establish the process for target and goal-setting within this framework. The first is the coastal stratum, which extends from the local Mean Lower Low Water elevation to elevation 5 feet above Mean Sea Level and is likely to be affected by frequent tidal flooding or inundation by 2070. The coastal stratum is the zone where emergent tidal wetland restoration would be conducted and includes low lying coastal uplands which serve as important tidal wetland buffers that will be critically important in the future as lands reserved to accommodate ideal wetland migration in response to sea level rise. The river floodplain stratum includes all hydrologically contiguous forested and non-forested wetlands within the river and stream corridors of the Tampa Bay watershed. Floodplain corridors provide vital watershed functions including fish and wildlife habitat and migratory pathways, floodwater attenuation and storage, erosion control, and delivery of complex organic matter to the estuarine food web. Finally, the upland stratum encompasses those areas outside of the coastal and river floodplain strata, including native upland habitats as well as hydrologically isolated wetlands. These habitats provide important aquifer recharge and wildlife habitat functions.

@@

@@

@@

The targets and goals (Table 6) identify where the 2017 CCMP Bay Habitat goals and strategies and the 2021-2025 Strategic Plan [@Burke20] thriving habitats and abundant wildlife programmatic priorities can be implemented. However, it also recognizes that the identified habitat protection and restoration areas will change over time, and will be revisited on a 10-year recurring cycle. A 30-year planning horizon (2050) is also identified based upon sea level rise projections developed specifically for Tampa Bay [@Burke19]. The coastal stratum (from the existing mean low water line to the approximate 5-foot contour) is projected to directly experience the effects of sea level rise by 2050, and is the primary focus area for coastal habitat protection and restoration activities. Land acquisition or protection (through conservation easements or other mechanisms) will be needed to ensure completion of targets and goals for both salt marsh and upland habitats (Fig 3).

Targets that maintain current coverage ("hold-the line strategies") were identified for habitats that appear to currently be sustained at acceptable levels. Evolving information such as an Oyster Habitat Suitability Index [@Boswell12] and ongoing mapping exercises will be used to identify optimal locations to conduct restoration activities that help achieve targets and goals. Coordination with the establishment of state-mandated minimum flows [@Munson07] will be necessary for restoration and maintenance of low-salinity salt marsh habitats that will experience higher salinities and rapid transition to mangroves under existing sea level rise scenarios [@Sherwood14; @Geselbracht15].

@@

The intertidal zone in Tampa Bay is currently experiencing dynamic change, driven by sea level rise and climate change, whereby mangrove forests are outcompeting salt marshes and salt barrens for the available niche space. Without increasing the total area of the intertidal zone, restoring a greater coverage of mangroves would reduce the niche space available for salt marshes and salt barrens. This phenomenon has been observed throughout the Gulf of Mexico, and has been attributed to both climate change (e.g., fewer freeze events) and sea level rise [@Comeaux12].

The observed gains in wetlands (Table 3b) are likely a reflection of: 1) the effectiveness of state and federal wetland regulatory programs; and 2) the cumulative gains resulting from, primarily, publicly-funded habitat restoration projects (Table 4). Minor gains in some emergent tidal wetlands (e.g., salt barrens) may also be a reflection of the landward expansion of the complex suite of these habitats associated with climate change and sea level rise. Gains in vegetated non-forested freshwater wetlands are related to the clearing of forested wetlands followed by the creation of herbaceous mitigation areas and stormwater systems.

@@

As discussed, land acquisition for coastal habitat restoration must prioritize adjacent low-lying coastal uplands to serve as buffers to accommodate future landward migration of tidal wetlands in response to sea level rise. Where public acquisition is not possible, other conservation mechanisms need to be explored. Coastal setbacks, buffers, or public easements are traditionally used to restrict development within a given distance from the shoreline. A rolling easement is a dynamic mechanism that "rolls" landward as sea levels rise and cause tidal encroachments onto low-lying coastal uplands [@Titus09]. The application of rolling easements in Tampa Bay could disincentivize more intense urban development (e.g., discourage up-zoning) of low-lying coastal uplands that may be currently in less intense agricultural or recreational (e.g., golf courses) land uses. Under a rolling easement, landowners would be able to maintain current economic uses, while "reserving" such lands to accommodate tidal wetland migration with advancing sea level rise.

>>>>

The health of estuarine systems and coastal habitats is inextricably linked to land uses and management of the watershed [@Yoskowitz15]. Coastal habitats provide multiple ecosystem services, including wildlife shelter and migratory corridors [@Yoskowitz15], fisheries production [@Houde93], water quality improvement [@Kushlan90; @Sprandel00; @AvilaGarcia20], erosion and flood attenuation [@Calil15; @Menendez18], carbon sequestration [@Dontis20] and recreation [@Chung18]. Pervasive stressors can negatively impact the services provided by these habitats. Anthropogenic land development can remove or substantially modify existing habitats, and can alter hydrology of nearby streams and rivers [@Theobald97; @Sim14]. Related changes in habitats in response to climate change (sea-level rise and warming) include landward migration of mangroves into salt marshes, upstream migration of salt marshes within tidal tributaries, and upland forest migration [@Brinson95; @Vogelmann12; @Cavanaugh19].

The combined effects of land development and climate change are especially problematic for prioritizing habitat restoration or conservation activities in coastal environments. Landward migration of critical habitats in response to sea-level rise may not be possible due to anthropogenic barriers in the watershed. Sea-level rise can occur quicker than landward migration of salt marshes and the upland slope may already be lost to urban development and hardening [@Titus09].  The coastal squeeze of development and climate change diminishes the available space for future restoration in urban estuaries, thereby impacting the variety of ecosystem services provided by these habitats and the wildlife they support [@Enwright15]. Given projected habitat losses without management intervention and the limited resources available (i.e., time, land, funding, and labor force), appropriate and realistic sites for restoration need to be identified that account for future stressors and past trends.

Past approaches for guiding restoration planning have been successfully used in other contexts, but they do not fully balance competing needs among public and private sectors, nor do they fully account for anticipated effects of multiple stressors. For example, an integrated watershed approach [@EPA96] has been utilized since the early 1990s to diagnose and manage water quantity and quality problems by addressing issues within hydrologically-defined geographic areas. Additionally, the habitat mosaic approach [@Henningsen05] of including multiple habitat types within restoration projects has been recognized as an effective means of allowing ecosystem state changes in response to different environmental pressures [@Duarte09; @Palmer09]. Adaptive management [@Holling78; @Gregory06] components have been increasingly used to address challenges of sea-level rise, climate change, and development stressors, including monitoring to identify critical restoration decision points and needed intervention with contingency plans. Elements of each of these approaches could be combined to create a more holistic approach to guide restoration and conservation activities for coastal habitats in urban settings.

The Tampa Bay watershed (Florida, USA) is a valuable case study for developing a habitat restoration and conservation plan that addresses pervasive coastal stressors. Compared to other estuaries, the ratio of watershed to estuary area is small and the area is heavily developed with 42% of land use classified as urban and suburban residential [@SWFWMD18]. A retrospective approach to setting habitat protection and restoration targets in Tampa Bay was previously employed [@Lewis96; @Robison10; @Cicchetti11; @Russell15]. Priority was given to restoration activities focused on habitat types that were important for a suite of estuarine faunal guilds disproportionately lost or degraded compared to a benchmark period. Criticisms of this approach included a lack of consideration for future sea-level rise and other climate change factors [@Yoskowitz15], use of expanded and different habitats outside the Tampa Bay watershed [@Robison10], lack of attention to upland or freshwater wetland habitats, and little recognition of land development trends or actual available space for restoration efforts.  These challenges are shared by restoration practitioners in other coastal environments and an approach that accommodates these challenges for planning would be highly transferable.

This paper describes an approach for habitat restoration and conservation planning that addresses the above challenges by considering the whole watershed, addressing historical changes, focusing on trajectories that have occurred during contemporary time periods, and considering both current and future stressors -- particularly land development and sea-level rise. Relatively consistent data are available for most Tampa Bay habitats of interest (1988 to 2018), representing a time period when federal, state and local regulations were in effect and regional impacts from climate change are documented [@Raabe12; @Cavanaugh14]. The approach establishes a framework that can guide both watershed-level habitat planning and site-level restoration activities and incorporates applicable elements of other habitat restoration paradigms discussed above [@Palmer09]. The general approach includes 1) designation of habitat types by strata relative to the aquatic-terrestrial gradient, 2) quantification of historical trends by habitat types to identify appropriate future targets in coverage, and 3) identification of opportunity areas that could be used by practitioners to achieve restoration goals based on habitat type and past trajectories.  The outcomes are spatially-specific by providing maps to identity opportunity areas and reproducible using an open science workflow [@Lowndes17] that allows regular updates as new data become available.

@@

Addressing the full suite of habitats in a watershed is critical for large-scale restoration planning efforts [@Palmer09; @Lamb18]**.** The major habitat types of Tampa Bay can be described and organized relative to tidal influence and location in the watershed. Subtidal habitats include those that are submerged all or most of the time; emergent tidal wetlands include those that are submerged during high tides but exposed during low tides; and supratidal habitats include those that occur above the high tide line.

@@

Three strata, distinct geographic breakpoints in the watershed where different habitat management and restoration activities could take place, were created to establish the process for target and goal-setting within this framework. The first is the coastal stratum, which extends from the local Mean Lower Low Water elevation to elevation 5 feet above Mean Sea Level and is likely to be affected by frequent tidal flooding or inundation by 2070. The coastal stratum is the zone where emergent tidal wetland restoration would be conducted and includes low lying coastal uplands which serve as important tidal wetland buffers that will be critically important in the future as lands reserved to accommodate ideal wetland migration in response to sea-level rise. The river floodplain stratum includes all hydrologically contiguous forested and non-forested wetlands within the river and stream corridors of the Tampa Bay watershed. Floodplain corridors provide vital watershed functions including fish and wildlife habitat and migratory pathways, floodwater attenuation and storage, erosion control, and delivery of complex organic matter to the estuarine food web. Finally, the upland stratum encompasses those areas outside of the coastal and river floodplain strata, including native upland habitats as well as hydrologically isolated wetlands. These habitats provide important aquifer recharge and wildlife habitat functions.

@@

Table 3

@@

Table 4

@@

The targets and goals (Table 6) identify where the 2017 CCMP Bay Habitat goals and strategies and the 2021-2025 Strategic Plan [@Burke20] thriving habitats and abundant wildlife programmatic priorities can be implemented. However, it also recognizes that the identified habitat protection and restoration areas will change over time, and will be revisited on a 10-year recurring cycle. A 30-year planning horizon (2050) is also identified based upon sea-level rise projections developed specifically for Tampa Bay [@Burke19]. The coastal stratum (from the existing mean low water line to the approximate 5-foot contour) is projected to directly experience the effects of sea-level rise by 2050, and is the primary focus area for coastal habitat protection and restoration activities. Land acquisition or protection (through conservation easements or other mechanisms) will be needed to ensure completion of targets and goals for both salt marsh and upland habitats (Fig 3).

Targets that maintain current coverage ("hold-the line strategies") were identified for habitats that appear to currently be sustained at acceptable levels. Evolving information such as an Oyster Habitat Suitability Index [@Boswell12] and ongoing mapping exercises will be used to identify optimal locations to conduct restoration activities that help achieve targets and goals. Coordination with the establishment of state-mandated minimum flows [@Munson07] will be necessary for restoration and maintenance of low-salinity salt marsh habitats that will experience higher salinities and rapid transition to mangroves under existing sea-level rise scenarios [@Sherwood14; @Geselbracht15].

@@

The intertidal zone in Tampa Bay is currently experiencing dynamic change, driven by sea-level rise and climate change, whereby mangrove forests are outcompeting salt marshes and salt barrens for the available niche space. Without increasing the total area of the intertidal zone, restoring a greater coverage of mangroves would reduce the niche space available for salt marshes and salt barrens. This phenomenon has been observed throughout the Gulf of Mexico, and has been attributed to both climate change (e.g., fewer freeze events) and sea-level rise [@Comeaux12].

The observed gains in wetlands (Table 3b) are likely a reflection of: 1) the effectiveness of state and federal wetland regulatory programs; and 2) the cumulative gains resulting from, primarily, publicly-funded habitat restoration projects (Table 4). Minor gains in some emergent tidal wetlands (e.g., salt barrens) may also be a reflection of the landward expansion of the complex suite of these habitats associated with climate change and sea-level rise. Gains in vegetated non-forested freshwater wetlands are related to the clearing of forested wetlands followed by the creation of herbaceous mitigation areas and stormwater systems.

@@

As discussed, land acquisition for coastal habitat restoration must prioritize adjacent low-lying coastal uplands to serve as buffers to accommodate future landward migration of tidal wetlands in response to sea-level rise. Where public acquisition is not possible, other conservation mechanisms need to be explored. Coastal setbacks, buffers, or public easements are traditionally used to restrict development within a given distance from the shoreline. A rolling easement is a dynamic mechanism that "rolls" landward as sea levels rise and cause tidal encroachments onto low-lying coastal uplands [@Titus09]. The application of rolling easements in Tampa Bay could disincentivize more intense urban development (e.g., discourage up-zoning) of low-lying coastal uplands that may be currently in less intense agricultural or recreational (e.g., golf courses) land uses. Under a rolling easement, landowners would be able to maintain current economic uses, while "reserving" such lands to accommodate tidal wetland migration with advancing sea-level rise.
